Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-03-15 paper/paper.tex
sum combination algorithm

\begin{equation}
\eqname{T$k$}
\begin{split}
    s \oplus t &= \texttt{largest}(s \cup t, k)\\
    s \odot t &= \texttt{largest}(\{a+b| a \in s, b\in t\}, k)\\
    \mymathbb{0} &= -\infty^{\otimes k}\\
    \mymathbb{1} &= -\infty^{\otimes k-1} \otimes 0
\end{split}\label{eq:ext-tropical}
\end{equation}
where $\texttt{largest}(s, k)$ means truncating the set by only keeping $k$ largest values in the set $s$.
The computation of $s \odot t$ is a maximum sum combination problem which can be done in time $\sim k\log(k)$.
The vertex tensor and edge tensor become
\begin{equation}
    W^{{\rm T}k}(-\infty^{\otimes k-1} \otimes 1, w_v) = \left(\begin{matrix}
        \mymathbb{1} \\
        -\infty^{\otimes k-1} \otimes w_v
    \end{matrix}\right),   
    \qquad
        B^{{\rm T}k} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right),
\end{equation}
where we have used the relation $(-\infty^{\otimes k-1} \otimes 1)^{w_v} = -\infty^{\otimes k-1} \otimes w_v$.
Similar to the tropical numbers, we can combine it with a bit string sampler (\Eq{eq:singleconfig}) for finding the configuration of each size.
Since the $\odot$ operation of configuration sampler is not used, the resulting configurations are deterministic and complete.

@@

>>>>

\begin{equation}
\eqname{T$k$}
\begin{split}
    s \oplus t &= \texttt{largest}(s \cup t, k)\\
    s \odot t &= \texttt{largest}(\{a+b| a \in s, b\in t\}, k)\\
    \mymathbb{0} &= -\infty^{\otimes k}\\
    \mymathbb{1} &= -\infty^{\otimes k-1} \otimes 0
\end{split}\label{eq:ext-tropical}
\end{equation}
where $\texttt{largest}(s, k)$ means truncating the set by only keeping $k$ largest values in the set $s$.
The computation of $s \odot t$ is a maximum sum combination problem which can be done in time $\sim k\log(k)$,
one can find an implementation in \App{sec:maxsum}.
The vertex tensor and edge tensor become
\begin{equation}
    W^{{\rm T}k}(-\infty^{\otimes k-1} \otimes 1, w_v) = \left(\begin{matrix}
        \mymathbb{1} \\
        -\infty^{\otimes k-1} \otimes w_v
    \end{matrix}\right),   
    \qquad
        B^{{\rm T}k} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right),
\end{equation}
where we have used the relation $(-\infty^{\otimes k-1} \otimes 1)^{w_v} = -\infty^{\otimes k-1} \otimes w_v$.
Similar to the tropical numbers, we can combine it with a bit string sampler (\Eq{eq:singleconfig}) for finding the configuration of each size.
Since the $\odot$ operation of configuration sampler is not used, the resulting configurations are deterministic and complete.

@@

\section{Computing maximum sum combination}\label{sec:maxsum}
Given two sets $A$ and $B$ of the same size $n$.
It is known that the maximum $n$ sum combination of $A$ and $B$ can be computed in time $n\log(n)$.
The standard approach to solve the sum combination problem requires storing the variables in a heap - a highly dynamic binary tree structure which can can be much slower to manipulate than arrays.
In the following, we show an algorithm that has roughly the same complexity (given the data range is not exponentially large)
but does not need a heap.
This algorithm first sorts both $A$ and $B$, then use the bisection to find the $n$-th largest value in the sum combination.
The key point is we can count the number of entries greater than a specific value
in the sum combination of $A$ and $B$ in linear time.
We summarize the algorithm as in Algorithm~\ref{alg:sumcombination}.

\LinesNumberedHidden
\begin{algorithm}[!ht]
    \SetKwProg{Fn}{function}{}{end}
    \small
    \SetAlgoNoLine
    Let $A$ and $B$ be two sets of size $n$\;

    \tcp{sort $A$ and $B$ in ascending order}
    $A \gets {\rm sort}(A)$\;

    $B \gets {\rm sort}(B)$\;

    \tcp{use bisection to find the $n$-th largest value in sum combination}

    ${\rm high} \gets A_n+B_n$\;

    ${\rm low} \gets A_1+B_n$\;

    \While{true}{
        ${\rm mid} \gets ({\rm high} + {\rm low}) / 2$\;

        $c \gets \texttt{count\_geq}(n, A, B, {\rm mid})$\;

        \uIf{$c > n$}{
            ${\rm low} \gets {\rm mid}$\;
        }\uElseIf{$c = n$}{
            \Return \texttt{collect\_geq}$(n, A, B, {\rm mid})$\;
        }\Else{
            ${\rm high} \gets {\rm mid}$\;
        }
    }

    \Fn{{\rm \texttt{count\_geq}}($n$, $A$, $B$, $v$)}{
        $k \gets 1$\; \tcp*[l]{number of entries in $A$ s.t. $a+b\geq v$}

        $a \gets A_{n}$\; \tcp*[l]{the smallest entry in $A$ s.t. $a+b\geq v$}

        $c \gets 0$\; \tcp*[l]{the counting of sum combinations s.t. $a+b\geq v$}

        \For{$q$ = $n, n-1 \ldots 1$}{
            $b \gets B_{n-q+1}$\;

            \While{$k < n$ \textbf{and} $a+b \geq v$}{
                $k \gets k+1$\;

                $a \gets A_{n-k+1}$\;
            }

            \uIf{$a+b \geq v$}{
                $c \gets c + k$\;
            }\Else{
                $c \gets c + k-1$\;
            }
        }
        \Return c\;
    }

    % \Fn{\rm collect\_geq($n$, $A$, $B$, $v$)}{
    %     $k \gets 1$\; \tcp*[l]{number of entries in $A$ s.t. $a+b\geq v$}

    %     $a \gets A_{n}$\; \tcp*[l]{the smallest entry in $A$ s.t. $a+b\geq v$}

    %     $S \gets \{\}$ \tcp*[l]{the sum combinations s.t. $a+b\geq v$}

    %     \For{$q$ = $n, n-1 \ldots 1$}{
    %         $b \gets B_{n-q+1}$\;

    %         \While{$k < n$ \textbf{and} $a+b \geq v$}{
    %             $k \gets k+1$\;

    %             $a \gets A_{n-k+1}$\;
    %         }

    %         \uIf{$a+b \geq v$}{
    %             $m \gets k$\;
    %         }\Else{
    %             $m \gets k-1$\;
    %         }
    %         \For{$j$ = \KwTo $1,2\ldots,m$}{
    %             $S \gets S \cup \{b + A_{n-j+1}\}$
    %         }
    %     }
    %     \Return S\;
    % }
    \caption{Fast sum combination without using heap}\label{alg:sumcombination} 
\end{algorithm}

Function \texttt{collect\_geq} is similar the \texttt{count\_geq} except the counting is replace by the item collection.
Note in function \texttt{count\_geq}, variable $k$ monotoneously increase while $q$ monotoneously decrease in each iteration,
the total number of iterations must be smaller or equal to $2n$.
Here for simplicity, we do not handle the $-\infty$s in $A$ and $B$ and the degeneracy in the sums.
They need to be taken seriously in practical implementations.
